\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-11 paper.qmd
Added 2D model explanation with equations

@@

@@

**(d) Assign NLDR data to hexagons**

@@

The mathematical expression for the hexagonal bin centers can be defined based on the hexagonal grid coordinates. Let's denote the hexagonal grid coordinates as \( (q, r) \), where \( q \) and \( r \) are integers representing the coordinates of the hexagon in the grid.

For a regular hexagonal grid, the coordinates of the hexagon centers can be calculated using the following formulas:
<!--
\[
x = \frac{3}{2} \times q \times s
\]
\[
y = \frac{\sqrt{3}}{2} \times q \times s + \sqrt{3} \times r \times s
\]-->

Where:
- \( s \) is the side length of the hexagon.
- \( q \) represents the column index of the hexagon in the grid.
- \( r \) represents the row index of the hexagon in the grid.
- \( x \) and \( y \) are the Cartesian coordinates of the center of the hexagon.

This formula calculates the \( x \) and \( y \) coordinates of the center of each hexagon based on its grid coordinates \( q \) and \( r \), allowing you to determine the position of each hexagon in the plane.

@@

Delaunay triangulation is a geometric algorithm used to create a triangular mesh from a set of points in a plane. The key principle behind Delaunay triangulation is to connect points in such a way that no point falls within the circumcircle of any triangle formed by the points.

Mathematically, given a set of points \( P \) in a plane, the Delaunay triangulation constructs a set of triangles such that:

1. Each triangle is formed by connecting three points from \( P \).
2. No point in \( P \) is inside the circumcircle of any triangle in the triangulation.

Formally, the Delaunay triangulation can be defined as follows:

Let \( P = \{p_1, p_2, ..., p_n\} \) be a set of \( n \) points in the plane. The Delaunay triangulation of \( P \), denoted as \( DT(P) \), is a triangulation of the convex hull of \( P \) such that the circumcircle of every triangle in the triangulation contains no other points from \( P \).

@@

$$
b = b_1 \times b_2
$${#eq-equation5}

**Step 2: Obtain bin centroids or bin means**

As a result of hexagonal binning for high-D data, all the high-D data are clustered into hexagons. In this step, the bin centroids ($C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2})$) (see @fig-meth Step 2) are obtained [@Carr2013].

**Step 3: Triangulate bin centroids**

In this step, the algorithm proceeds to triangulate the hexagonal bin centroids (see @fig-meth Step 3). Triangulation is a fundamental process in computational geometry and computer graphics that involves dividing a set of points in a given space into interconnected triangles [@article30]. One common algorithm used for triangulation is Delaunay triangulation [@article26, @article54], where points are connected in a way that maximizes the minimum angles of the resulting triangles, leading to a more regular and well-conditioned triangulation.

Since we are working with the centroids of regular hexagonal bins, the resulting mesh will predominantly comprise equal-sized regular triangles. However, the triangulation also helps span any gaps that may exist between clusters of points, allowing for a more complete and interconnected representation of the data.

@@

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_components=2, n_neighbors=10, init=random, MN_ratio=0.9, FP_ratio=2.0) applied for the digit 1 of the MNIST dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with PaCMAP (<https://youtu.be/CNXjWZMyh_M>)"
#| label: fig-nldervisMNISTPACMAP
#| out-width: 100%
#| fig-pos: h

>>>>

The calculation of $y_{max}$ involves several steps. First, the aspect ratio ($ar$) is computed by dividing the range of the second 2D embedding component ($r_2$) by the range of the first 2D embedding component ($r_1$) (see @eq-equation1). Then, the hexagon ratio ($hr$) is determined by dividing the height of the hexagon ($hb$) by its width ($wb$) (see @eq-equation2). Finally, $y_{max}$ is derived by taking the ceiling of $\frac{ar}{hr}$ and multiplying it by $hr$. This process ensures that $y_{max}$ is an integer multiple of $hr$, accommodating the grid layout of the hexagonal bins.

@@

Hexagonating NLDR data (see @fig-meth Step 2) involves partitioning the 2D embedding data into hexagonal bins, a technique commonly referred to as hexagonal binning [@Carr1987, @article66]. This method use a hexagonal grid to create a bivariate histogram that effectively visualizes the structure of high-D data. Hexagons, one of only three regular polygons capable of tessellating a plane [@Carr2013], offer unique advantages due to their symmetry of nearest neighbors and maximal number of sides for such tessellations [@Dan2023]. This geometric property makes hexagons more efficient in covering the plane compared to other regular tessellations and reduces visual bias when displaying data densities [@Dan2023]. In our algorithm, we aim to conduct regular hexagonal binning (**cite quollr paper**), involving the computation of hexagonal grid configurations, the generation of the grid, and the assignment of the 2D embedding to hexagons.

<!--
**(a) Determine the number of bins along the x axis** ($b_1$)

@@

**(d) Assign NLDR data to hexagons**-->

@@

In the previous step, the algorithm clusters the 2D embedding data into hexagons. Following this, in this step, the bin centroids or bin means (see @fig-meth Step 3) are obtained [@Carr2013].

The bin centroid ($C_k^{(2)}$) for a $k^{th}$ hexagon with hexagonal grid coordinates $(h^{k}x_{i}, h^{k}y_{i})$, where $i = 1 \dots 6$ can be defined as:

$$
C_k^{(2)} = (C_{ky_1}, C_{ky_2}) = \left(\frac{\sum_{i=1}^{6} h^{k}x_{i}}{6}, \frac{\sum_{i=1}^{6} h^{k}y_{i}}{6}\right).
$$ {#eq-equation4}

Also, the bin mean ($C_k^{(2)}$) is defined as the mean of the data points within the $k^{th}$ hexagon (see @eq-equation5).

$$
C_k^{(2)} = (C_{ky_1}, C_{ky_2}) = \left(\frac{1}{n_k} \sum_{i=1}^{n_k} y_{1i}, \frac{1}{n_k} \sum_{i=1}^{n_k} y_{2i}\right),
$$ {#eq-equation5}

where $n_k$ is the number of data points within the hexagon, $y_{1i}$ and $y_{2i}$ are the $x$ and $y$ coordinates of the $i^{th}$ data point within the hexagon.

@@

In this step, the algorithm proceeds to triangulate the hexagonal bin centroids or bin means (see @fig-meth Step 4). Triangulation is a fundamental process in computational geometry and computer graphics that involves dividing a set of points in a given space into interconnected triangles [@article30]. One common algorithm used for triangulation is Delaunay triangulation [@article26, @article54], where points are connected in a way that maximizes the minimum angles of the resulting triangles, leading to a more regular and well-conditioned triangulation.

Delaunay triangulation can be defined as follows:

Let $C^{(2)} = \{C_1^{(2)}, C_2^{(2)}, ..., C_m^{(2)}\}$ be a set of $m$ bin centroids or bin means in the plane. Delaunay triangulation of $C^{(2)}$, denoted as $DT(C^{(2)})$, is a triangulation of the convex hull of $C^{(2)}$ such that the circumcircle of every triangle in the triangulation contains no other points from $C^{(2)}$.

Given that the hexagons are regular, the resulting triangles will mostly be equilateral.

@@

$$
b = b_1 \times b_2
$${#eq-equation5}
-->

@@

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_components=2, n_neighbors=10, init=random, MN_ratio=0.9, FP_ratio=2.0) applied for the digit 1 of the MNIST dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with PaCMAP ()"
#| label: fig-nldervisMNISTPACMAP
#| out-width: 100%
#| fig-pos: h

### 2024-03-11 paper.tex
Added 2D model explanation with equations

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, breakable, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi

@@

@@

\textbf{(a) Determine the number of bins along the x axis} (\(b_1\))

\textbf{(b) Determine the number of bins along the y axis} (\(b_2\))

\textbf{(c) Determine the total number of bins} (\(b\))

\textbf{(d) Assign NLDR data to hexagons}

@@

The mathematical expression for the hexagonal bin centers can be defined
based on the hexagonal grid coordinates. Let's denote the hexagonal grid
coordinates as ( (q, r) ), where ( q ) and ( r ) are integers
representing the coordinates of the hexagon in the grid.

For a regular hexagonal grid, the coordinates of the hexagon centers can
be calculated using the following formulas:

Where: - ( s ) is the side length of the hexagon. - ( q ) represents the
column index of the hexagon in the grid. - ( r ) represents the row
index of the hexagon in the grid. - ( x ) and ( y ) are the Cartesian
coordinates of the center of the hexagon.

This formula calculates the ( x ) and ( y ) coordinates of the center of
each hexagon based on its grid coordinates ( q ) and ( r ), allowing you
to determine the position of each hexagon in the plane.

@@

Delaunay triangulation is a geometric algorithm used to create a
triangular mesh from a set of points in a plane. The key principle
behind Delaunay triangulation is to connect points in such a way that no
point falls within the circumcircle of any triangle formed by the
points.

Mathematically, given a set of points ( P ) in a plane, the Delaunay
triangulation constructs a set of triangles such that:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Each triangle is formed by connecting three points from ( P ).
\item
  No point in ( P ) is inside the circumcircle of any triangle in the
  triangulation.
\end{enumerate}

Formally, the Delaunay triangulation can be defined as follows:

Let ( P = \{p\_1, p\_2, \ldots, p\_n\} ) be a set of ( n ) points in the
plane. The Delaunay triangulation of ( P ), denoted as ( DT(P) ), is a
triangulation of the convex hull of ( P ) such that the circumcircle of
every triangle in the triangulation contains no other points from ( P ).

\textbf{Step 1: Computing the hexagonal grid configuration}

The hexagonal grid, formed through hexagonal binning
\citep[\citet{article66}]{Carr1987}, serves as a type of bivariate
histogram employed to visualize the structure of high-D data. Hexagons,
being one of only three regular polygons capable of tessellating a plane
\citep{Carr2013}, possess both symmetry of nearest neighbors and the
maximum number of sides for a regular tessellation of the plane. This
unique combination makes hexagons more efficient in covering the plane
compared to other regular tessellations. Additionally, hexagons exhibit
lower visual bias when displaying densities, setting them apart from
other regular tessellations \citep{Dan2023}. In our algorithm, hexagonal
binning is used as the initial step of constructing the 2D model and the
total number of bins (\(b\)) is the crucial parameter that defines the
granularity of the hexagonal grid.

\textbf{(a) Determine the number of bins along the x axis} (\(b_1\))

First, the number of bins along the x-axis (\(b_1\)) is computed using
the relationship between the diameter (\(h\)) and the area (\(A\)) of
regular hexagons (see Equation~\ref{eq-equation3}).

In the process of computing hexagonal bin configurations, the first step
involves determining the number of bins along the x-axis. This begins
with the initialization of parameters such as the hexagonal size (\(s\))
and the buffer along the x-axis. Subsequently, the range of the first 2D
embedding component (\(r_1\)) is calculated to ensure comprehensive
coverage of the data. This range is then adjusted by incorporating the
buffer amount, effectively expanding the boundary to accommodate
potential outliers or edge cases. The default buffer amount along the
x-axis is \(\sqrt{3} * s * 1.5\). Next, the horizontal spacing (\(h\))
between hexagons is computed based on the hexagon size. Finally, the
number of bins along the x-axis is computed based on the adjusted range
(\(r_1 + buffer_{x}\)) and the horizontal spacing (\(h\)).

\begin{equation}\protect\hypertarget{eq-equation3}{}{
 h = \sqrt{3} * s
}\label{eq-equation3}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation4}{}{
 b_1 = \frac{r_1 + buffer_{x}}{h}
}\label{eq-equation4}\end{equation}

\textbf{(b) Determine the number of bins along the y-axis} (\(b_2\))

Next, the number of bins along the y-axis is computed based on the
number of bins along the x-axis (\(b_1\)) and the shape parameter
(\(s\)) (see Equation~\ref{eq-equation4}) \citep{Carr2013}.

\begin{equation}\protect\hypertarget{eq-equation5}{}{
  v = 1.5 * s
}\label{eq-equation5}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation6}{}{
 b_2 = \frac{r_2 + buffer_{y}}{v}
}\label{eq-equation6}\end{equation}

\textbf{(c) Determine the total number of bins} (\(b\))

The total number of bins is determined by multiplying the number of bins
along the x-axis (\(b_1\)) with the number of bins along the y-axis
(\(b_2\)) (see Equation~\ref{eq-equation5}).

\begin{equation}\protect\hypertarget{eq-equation5}{}{
b = b_1 \times b_2
}\label{eq-equation5}\end{equation}

\textbf{Step 2: Obtain bin centroids or bin means}

As a result of hexagonal binning for high-D data, all the high-D data
are clustered into hexagons. In this step, the bin centroids
(\(C_k^{(2)} \equiv (C_{ky_1}, C_{ky_2})\)) (see Figure~\ref{fig-meth}
Step 2) are obtained \citep{Carr2013}.

\textbf{Step 3: Triangulate bin centroids}

In this step, the algorithm proceeds to triangulate the hexagonal bin
centroids (see Figure~\ref{fig-meth} Step 3). Triangulation is a
fundamental process in computational geometry and computer graphics that
involves dividing a set of points in a given space into interconnected
triangles \citep{article30}. One common algorithm used for triangulation
is Delaunay triangulation \citep[\citet{article54}]{article26}, where
points are connected in a way that maximizes the minimum angles of the
resulting triangles, leading to a more regular and well-conditioned
triangulation.

Since we are working with the centroids of regular hexagonal bins, the
resulting mesh will predominantly comprise equal-sized regular
triangles. However, the triangulation also helps span any gaps that may
exist between clusters of points, allowing for a more complete and
interconnected representation of the data.

@@

\caption{\label{fig-nldervisMNISTPACMAP}(a) 2D layout from UMAP
(n\_components=2, n\_neighbors=10, init=random, MN\_ratio=0.9,
FP\_ratio=2.0) applied for the digit 1 of the MNIST dataset. Is this a
best representation of the original data?, (b) Model in the 2D space
with PaCMAP (\url{https://youtu.be/CNXjWZMyh_M})}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, breakable]}{\end{tcolorbox}}\fi

@@

The calculation of \(y_{max}\) involves several steps. First, the aspect
ratio (\(ar\)) is computed by dividing the range of the second 2D
embedding component (\(r_2\)) by the range of the first 2D embedding
component (\(r_1\)) (see Equation~\ref{eq-equation1}). Then, the hexagon
ratio (\(hr\)) is determined by dividing the height of the hexagon
(\(hb\)) by its width (\(wb\)) (see Equation~\ref{eq-equation2}).
Finally, \(y_{max}\) is derived by taking the ceiling of
\(\frac{ar}{hr}\) and multiplying it by \(hr\). This process ensures
that \(y_{max}\) is an integer multiple of \(hr\), accommodating the
grid layout of the hexagonal bins.

@@

Hexagonating NLDR data (see Figure~\ref{fig-meth} Step 2) involves
partitioning the 2D embedding data into hexagonal bins, a technique
commonly referred to as hexagonal binning
\citep[\citet{article66}]{Carr1987}. This method use a hexagonal grid to
create a bivariate histogram that effectively visualizes the structure
of high-D data. Hexagons, one of only three regular polygons capable of
tessellating a plane \citep{Carr2013}, offer unique advantages due to
their symmetry of nearest neighbors and maximal number of sides for such
tessellations \citep{Dan2023}. This geometric property makes hexagons
more efficient in covering the plane compared to other regular
tessellations and reduces visual bias when displaying data densities
\citep{Dan2023}. In our algorithm, we aim to conduct regular hexagonal
binning (\textbf{cite quollr paper}), involving the computation of
hexagonal grid configurations, the generation of the grid, and the
assignment of the 2D embedding to hexagons.

@@

In the previous step, the algorithm clusters the 2D embedding data into
hexagons. Following this, in this step, the bin centroids or bin means
(see Figure~\ref{fig-meth} Step 3) are obtained \citep{Carr2013}.

The bin centroid (\(C_k^{(2)}\)) for a \(k^{th}\) hexagon with hexagonal
grid coordinates \((h^{k}x_{i}, h^{k}y_{i})\), where \(i = 1 \dots 6\)
can be defined as:

\begin{equation}\protect\hypertarget{eq-equation4}{}{
C_k^{(2)} = (C_{ky_1}, C_{ky_2}) = \left(\frac{\sum_{i=1}^{6} h^{k}x_{i}}{6}, \frac{\sum_{i=1}^{6} h^{k}y_{i}}{6}\right).
}\label{eq-equation4}\end{equation}

Also, the bin mean (\(C_k^{(2)}\)) is defined as the mean of the data
points within the \(k^{th}\) hexagon (see Equation~\ref{eq-equation5}).

\begin{equation}\protect\hypertarget{eq-equation5}{}{
C_k^{(2)} = (C_{ky_1}, C_{ky_2}) = \left(\frac{1}{n_k} \sum_{i=1}^{n_k} y_{1i}, \frac{1}{n_k} \sum_{i=1}^{n_k} y_{2i}\right),
}\label{eq-equation5}\end{equation}

where \(n_k\) is the number of data points within the hexagon,
\(y_{1i}\) and \(y_{2i}\) are the \(x\) and \(y\) coordinates of the
\(i^{th}\) data point within the hexagon.

@@

In this step, the algorithm proceeds to triangulate the hexagonal bin
centroids or bin means (see Figure~\ref{fig-meth} Step 4). Triangulation
is a fundamental process in computational geometry and computer graphics
that involves dividing a set of points in a given space into
interconnected triangles \citep{article30}. One common algorithm used
for triangulation is Delaunay triangulation
\citep[\citet{article54}]{article26}, where points are connected in a
way that maximizes the minimum angles of the resulting triangles,
leading to a more regular and well-conditioned triangulation.

Delaunay triangulation can be defined as follows:

Let \(C^{(2)} = \{C_1^{(2)}, C_2^{(2)}, ..., C_m^{(2)}\}\) be a set of
\(m\) bin centroids or bin means in the plane. Delaunay triangulation of
\(C^{(2)}\), denoted as \(DT(C^{(2)})\), is a triangulation of the
convex hull of \(C^{(2)}\) such that the circumcircle of every triangle
in the triangulation contains no other points from \(C^{(2)}\).

Given that the hexagons are regular, the resulting triangles will mostly
be equilateral.

@@

\caption{\label{fig-nldervisMNISTPACMAP}(a) 2D layout from UMAP
(n\_components=2, n\_neighbors=10, init=random, MN\_ratio=0.9,
FP\_ratio=2.0) applied for the digit 1 of the MNIST dataset. Is this a
best representation of the original data?, (b) Model in the 2D space
with PaCMAP ()}
